Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-13 submission/manuscript.Rmd
edit results 5

While many germline genes act in both the male and female germline, some display sex-biased expression or have functions unique to eggs and sperm. To comprehensively assess whether germline gene sex corresponds with _Kdm5c_ mutant sex, we filtered our list of germline-enriched genes for egg (XX) and sperm (XY) biased genes. We defined sex-biased genes as those whose expression in the opposite sex, at any time point, is no greater than 20% of the gene's maximum expression in a given sex. This yielded `r nrow(subset(germgenes, sexBias == "XX"))` egg-biased, `r nrow(subset(germgenes, sexBias == "XY"))` sperm-biased, and `r nrow(subset(germgenes, sexBias == "unbiased"))` unbiased germline-enriched genes (Figure 4H), which is consistent with the testis overall having a more unique transcriptome than the ovary[@liComprehensiveMouseTranscriptomic2017]. We found egg, sperm, and unbiased germline genes were dyregulated in all _Kdm5c_ mutants (Figure 4I-L). Germline genes dysregulated exclusively in males and females were also not biased towards their corresponding germ cell sex. This indicates differences between male and female germline gene dysregulation is not due to sex-specific activation of sperm or egg-specific transcriptional programs. These results demonstrate that sex influences the degree of germline gene misexpression with KDM5C loss, but not the germ cell sex from which the germline genes originate.

@@

Previous work suggests KDM5C represses germline genes during early development, as re-expression of KDM5C in knockout neuronal cultures fails to suppress two hippocampal germline DEGs[@scandagliaLossKdm5cCauses2017]. KDM5C binds to several germline gene promoters in embryonic stem cells (ESCs) but is lost in neurons[@scandagliaLossKdm5cCauses2017]. However, the lack of a comprehensive list of germline-enriched genes prohibited systematic characterization of KDM5C binding at germline gene promoters. Thus, it is currently unclear if KDM5C is enriched at germline gene promoters, what types of germline genes KDM5C regulates, and if its binding is maintained at any germline genes in neurons.

@@

To address these questions, we analyzed KDM5C chromatin immunoprecipitation followed by DNA sequencing (ChIP-seq) datasets in EpiLCs[@samantaActivationXistEvolutionarily2022] and primary forebrain neuron cultures (PNCs)[@iwaseMouseModelXlinked2016]. EpiLCs had a higher total number of KDM5C peaks than PNCs (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC)`, PNCs: `r nrow(KDM5C_ChIP_PNC)`, MACS2 q < 0.1 and fold enrichment > 1, removal of _Kdm5c_-KO false positive peaks) and KDM5C was primarily localized to gene promoters in both cell types (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC_promo)`, PNCs: `r nrow(KDM5C_ChIP_PNC_promo)` +/- `r promorange`bp from the TSS), although PNCs showed increased localization to non-promoter regions (Figure 4A).

The majority of promoters bound by KDM5C in PNCs were also bound in EpiLCs (513 shared promoters), however a large portion of gene promoters were only bound by KDM5C in EpiLCs (3677 EpiLC only promoters) (Figure 4B). We then performed gene ontology analysis to compare the known functions of genes with KDM5C bound to their promoter in EpiLCs and PNCs. While there were no ontologies significantly enriched for genes only bound by KDM5C in PNCs, gene ontologies for genes shared between PNCs and EpiLCs were enriched for functions involving nucleic acid turnover, such as deoxyribonucleotide metabolic process (GO:0009262, p.adjust = `r signif(getGO("deoxyribonucleotide metabolic process", GO_KDM5C_ChIP, 'p.adjust', "Shared"), 3)`) (Figure 4C). Germline-specific ontologies were only enriched in promoters unique to EpiLCs, such as meiotic nuclear division (GO: 0007127 p.adjust = `r signif(getGO("meiosis I", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) and meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) (Figure 3C). We then evaluated KDM5C binding around the transcription start site (TSS) of all germline-enriched genes. In EpiLCs, We observed modest KDM5C signal at about half of all germline genes (Figure 4D). Based on our ChIP-seq peak criteria, KDM5C was strongly bound to about 10% of germline gene promoters in EpiLCs (Figure 4E). In condordance with our gene ontology results, we did not observe KDM5C accumiulation at any germline gene promtoers in PNCs (Figure 4D). Together, these results demonstrate KDM5C is significantly enriched at a subset of germline gene promoters in EpiLCs, including meiotic genes, but does not directly regulate germline genes in neurons.

Many germline-specific genes, including meiotic genes, are suppressed by the transcription factor heterodimers E2F6/DP1 and MGA/MAX, which respectively bind E2F and E-box motifs[@hurlinMgaDualspecificityTranscription1999; @pohlersRoleE2F6Restriction2005; @velascoDnmt3bRecruitmentE2F62010; @stielowMGAL3MBTL2E2F62018; @dahletE2F6InitiatesStable2021]. Thus, we identified transcription factor motifs enriched at KDM5C-bound or unbound germline gene promoters using HOMER[@heinzSimpleCombinationsLineageDetermining2010] (TSS +/- 500 bp, q-value < 0.1). MAX and E2F6 binding sites were significantly enriched at germline genes bound by KDM5C in EpiLCs, but not at germline genes unbound by KDM5C (MAX q-value: , E2F6 q-value:, E2F q-value: ) (Figure 4). One third of KDM5C-bound promoters contained the consensus sequence for either E2F6 (E2F, 5'-TCCCGC-3'), MGA (E-box, 5'-CACGTG-3'), or both, but only 17% of KDM5C-unbound genes contained these motifs (Figure 4). KDM5C-unbound germline genes were intstead enriched for multiple RFX transcription factor binding sites. RFX transcription factors bind X-box motifs[@gajiwalaStructureWingedhelixProtein2000] to promote ciliogenesis[@swobodaRFXTypeTranscriptionFactor2000; @ashiqueRfx4TranscriptionFactor2009]. Enriched RFX trancription factors included RFX2, a central regulator of post-meiotic spermatogeneis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016]. Interestingly, RFX2 mRNA is upregualted in _Kdm5c_-KO EpiLCs, but is also not a direct target of KDM5C (Figure 4). Thus, distinct transcription factor programs regulate KDM5C-bound and unbound germline genes.

Finally, we compared KDM5C binding at promoters of RNA-seq DEGs to determine if the germline mRNAs de-repressed in _Kdm5c_-KO EpiLCs and brain are direct targets of KDM5C (Figure 4F). In EpiLCs, KDM5C was bound to about one third of EpiLC-specific and brain-specific germline DEGs (EpiLC only: `r 18/50*100`%, Brain only: `r signif(7/21*100,3)`%). Again, we did not observe any KDM5C binding at germline gene promoters in PNCs, even for brain-specific DEGs (Figure 4H). Some notable differences in KDM5C binding for EpiLC-specific DEGs included _Dazl_ and _Stra8_, two key drivers of germline identity discussed above. Although both mRNAs are expressed in _Kdm5c_-KO EpiLCs, KDM5C is only bound to _Dazl_'s promoter and not _Stra8_'s in EpiLCs (Figure 4F). In contrast to the unique DEGs, 3 out of the 4 genes dysregulated in both the brain and EpiLCs are bound by KDM5C around the TSS (Figure 4G). Altogether, this suggests KDM5C decommssions germline genes in EpiLCs but is no longer recruited to germline gene promoters in neurons. Furthermore, the majority of germline mRNAs expressed in _Kdm5c_-KO cells are dysregulated independent of direct KDM5C recruitment to their promoters.

>>>>

While many germline genes act in both the male and female germline, some display sex-biased expression or have functions unique to eggs and sperm. To comprehensively assess whether germline gene sex corresponds with _Kdm5c_ mutant sex, we filtered our list of germline-enriched genes for egg (XX) and sperm (XY) biased genes. We defined sex-biased genes as those whose expression in the opposite sex, at any time point, is no greater than 20% of the gene's maximum expression in a given sex. This yielded `r nrow(subset(germgenes, sexBias == "XX"))` egg-biased, `r nrow(subset(germgenes, sexBias == "XY"))` sperm-biased, and `r nrow(subset(germgenes, sexBias == "unbiased"))` unbiased germline-enriched genes (Figure 4H), which is consistent with the testis overall having a more unique transcriptome than the ovary[@liComprehensiveMouseTranscriptomic2017]. We found egg, sperm, and unbiased germline genes were dyregulated in all _Kdm5c_ mutants (Figure 4I-J). Germline genes dysregulated exclusively in males and females were also not biased towards their corresponding germ cell sex. This indicates differences between male and female germline gene dysregulation is not due to sex-specific activation of sperm or egg-specific transcriptional programs. These results demonstrate that sex influences the degree of germline gene misexpression with KDM5C loss, but not the germ cell sex from which the germline genes originate.

@@

Previous work suggests KDM5C represses germline genes during early development, as re-expression of KDM5C in knockout neuronal cultures fails to suppress two hippocampal germline DEGs[@scandagliaLossKdm5cCauses2017]. KDM5C binds to several germline gene promoters in embryonic stem cells (ESCs) but is lost in neurons[@scandagliaLossKdm5cCauses2017]. However, the lack of a comprehensive list of germline-enriched genes prohibited systematic characterization of KDM5C binding at germline gene promoters. It is currently unclear if KDM5C is enriched at germline gene promoters, what types of germline genes KDM5C regulates, and if its binding is maintained at any germline genes in neurons.

@@

To address these questions, we analyzed KDM5C chromatin immunoprecipitation followed by DNA sequencing (ChIP-seq) datasets in EpiLCs[@samantaActivationXistEvolutionarily2022] and primary forebrain neuron cultures (PNCs)[@iwaseMouseModelXlinked2016]. EpiLCs had a higher total number of KDM5C peaks than PNCs (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC)`, PNCs: `r nrow(KDM5C_ChIP_PNC)`, MACS2 q < 0.1 and fold enrichment > 1, removal of _Kdm5c_-KO false positive peaks) and KDM5C was primarily localized to gene promoters in both cell types (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC_promo)`, PNCs: `r nrow(KDM5C_ChIP_PNC_promo)` +/- `r promorange`bp from the TSS), although PNCs showed increased localization to non-promoter regions (Figure 5A).

The majority of promoters bound by KDM5C in PNCs were also bound in EpiLCs (513 shared promoters), however a large portion of gene promoters were only bound by KDM5C in EpiLCs (3677 EpiLC only promoters) (Figure 5B). Genes bound by KDM5C in both PNCs and EpiLCs were enriched for functions involving nucleic acid turnover, such as deoxyribonucleotide metabolic process (GO:0009262, p.adjust = `r signif(getGO("deoxyribonucleotide metabolic process", GO_KDM5C_ChIP, 'p.adjust', "Shared"), 3)`) (Figure 5C). Germline-specific ontologies were only enriched in promoters unique to EpiLCs, such as meiotic nuclear division (GO: 0007127 p.adjust = `r signif(getGO("meiosis I", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) and meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) (Figure 5C). There were no ontologies significantly enriched for genes only bound by KDM5C in PNCs. We next evaluated KDM5C binding around the transcription start site (TSS) of all germline-enriched genes. In EpiLCs, We observed modest KDM5C signal at about half of all germline genes (Figure 5D). Based on our ChIP-seq peak cut-off criteria, KDM5C was strongly bound to about 10% of germline gene promoters in EpiLCs (Figure 5E). In condordance with our gene ontology results, we did not observe KDM5C accumiulation at any germline gene promtoers in PNCs (Figure 5D). Together, these results demonstrate KDM5C is significantly enriched at a subset of germline gene promoters in EpiLCs, including meiotic genes, but does not directly regulate germline genes in neurons.

Many germline-specific genes are suppressed by the transcription factor heterodimers E2F6/DP1 and MGA/MAX, which respectively bind E2F and E-box motifs[@hurlinMgaDualspecificityTranscription1999; @pohlersRoleE2F6Restriction2005; @velascoDnmt3bRecruitmentE2F62010; @stielowMGAL3MBTL2E2F62018; @dahletE2F6InitiatesStable2021]. To elucidate if KDM5C is recruited to germline gene promoters by a similar mechanism, we used HOMER to identify transcription factor motifs enriched at KDM5C-bound or unbound germline gene promoters[@heinzSimpleCombinationsLineageDetermining2010] (TSS +/- 500 bp, q-value < 0.1). MAX and E2F6 binding sites were significantly enriched at germline genes bound by KDM5C in EpiLCs, but not at germline genes unbound by KDM5C (MAX q-value: 0.0068, E2F6 q-value: 0.0673, E2F q-value: 0.0917) (Figure 5F). One third of KDM5C-bound promoters contained the consensus sequence for either E2F6 (E2F, 5'-TCCCGC-3'), MGA (E-box, 5'-CACGTG-3'), or both, but only 17% of KDM5C-unbound genes contained these motifs (Figure 5G). KDM5C-unbound germline genes were intstead enriched for multiple RFX transcription factor binding sites (RFX q-value < 0.0001, Rfx2 q-value < 0.0001, Rfx5 q-value < 0.0001) (Figure 5H). RFX transcription factors bind X-box motifs[@gajiwalaStructureWingedhelixProtein2000] to promote ciliogenesis[@swobodaRFXTypeTranscriptionFactor2000; @ashiqueRfx4TranscriptionFactor2009]. Enriched RFX trancription factors included RFX2, a central regulator of post-meiotic spermatogeneis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016]. Interestingly, RFX2 mRNA is upregualted in _Kdm5c_-KO EpiLCs (Figure 5I), but is not a direct target of KDM5C. Thus, distinct transcription factor programs regulate the expression of KDM5C-bound and unbound germline genes.

Finally, we compared KDM5C binding at the promoters of RNA-seq DEGs to determine if the germline genes transcribed in _Kdm5c_-KO cells are direct targets of KDM5C (Figure 5J-L). In EpiLCs, KDM5C was bound to about one third of EpiLC-specific and brain-specific germline DEG promoters (EpiLC only: `r 18/50*100`%, Brain only: `r signif(7/21*100,3)`%). Again, we did not observe any KDM5C binding at germline gene promoters in PNCs, even for brain-specific DEGs. Some notable differences in KDM5C binding for EpiLC-specific DEGs included _Dazl_ and _Stra8_, two key drivers of germline identity discussed above. Although both mRNAs are expressed in _Kdm5c_-KO EpiLCs, KDM5C is only bound to _Dazl_'s promoter and not _Stra8_'s in EpiLCs (Figure 5K). In contrast to the unique DEGs, 3 out of the 4 genes dysregulated in both the brain and EpiLCs are bound by KDM5C around the TSS. Altogether, this suggests KDM5C decommssions a subset of germline genes in EpiLCs but is no longer recruited to germline gene promoters in neurons. Furthermore, the majority of germline mRNAs expressed in _Kdm5c_-KO cells are dysregulated independent of direct KDM5C recruitment to their promoters.
